Penalized regression for left-truncated and right-censored survival data

### 2021-01-28 manuscript.tex
update LT section and patient follow-up

\subsection{Left truncation}
In the absence of left truncation, the risk set $R_i = \{ j : y_j \geq t_i \}$ assumes that every individual is at risk of an event at his or her respective time 0 and continues to be at risk until the event occurs or the individual is censored. In this sense, all individuals enter the risk set at the same time 0 and leave only at death or censoring. However, when survival times are left-truncated, individuals are truly at risk prior to entering the study (e.g. prior to being observed for follow-up). Take \autoref{fig:followup} as an example.

\begin{figure}[h]
\centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/followup.pdf} 
\caption{Left-truncated and right-censored patient follow-up in a hypothetical study cohort, ordered chronologically by event time. Patients become eligible to enter the study after diagnosis (closed circle) and are followed until death (open circle) or censoring (cross). However, patients who die before the start of the study (dashed blue line) are left-truncated (in red), and only those who have survived until the study start (in black) are observed.}
\label{fig:followup}
\end{figure}

Survival is defined as the time from diagnosis until death, but the study does not begin until 2011 (dashed blue vertical line). Patients 1 and 2 (in red) are excluded from the analysis because of death and right censoring, respectively, prior to study eligibility. A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors'' (those who have survived a long time, such as patients 3 and 7) and the recently-diagnosed.

Fortunately, the partial likelihood can be easily modified to make $R_i$ the set of individuals who are at risk (have entered the study and are not censored or failed) at that time assuming quasi-independence between left truncation and failure time \cite{tsai1987note}. Let individual $i$ have survival data of the form $(v_i, y_i, e_i, x_i)$, where individual $i$ enters the study at time $v_i$. The left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$. In other words, subjects are only considered at risk in the analysis after they have been observed to be at risk (i.e. exceeded their left truncation time). For example, the at-risk time for patients in Figure 1 is counted only after the vertical blue line marking study entry.

@@

\subsection{Data generating process}
We generated $n$ patients and $p$ predictors to construct a $n \times p$ matrix, $X$. The first 11 predictors were from the CGDB and the remaining $\tilde{p}$ variables were simulated binary variables. The CGDB variables were the same $11$ variables used in the small $p$ model in \autoref{sec:rwd} and generated by randomly sampling rows of the original data with replacement. The matrix of binary variables, $\tilde{X}$ was simulated using a (latent) multivariate probit approach

>>>>

\subsection{Left truncation}
In the absence of left truncation, the risk set $R_i = \{ j : y_j \geq t_i \}$ assumes that every individual is at risk of an event at his or her respective time 0 and continues to be at risk until the event occurs or the individual is censored. In this sense, all individuals enter the risk set at the same time 0 and leave only at death or censoring. However, when survival times are left-truncated, individuals are truly at risk prior to entering the study cohort (e.g. prior to being observed for follow-up). Take \autoref{fig:followup} as an example.

\begin{figure}[h]
\centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/followup.pdf} 
\caption{Left-truncated and right-censored patient follow-up in a hypothetical study cohort, ordered chronologically by event time. Patients who receive a diagnosis (closed circle) become eligible to enter the cohort after reaching a milestone (black triangle), for example a genomic test. Patients are followed until death (open circle) or censoring (cross). However, patients who die before reaching the milestone are left-truncated (in red), and only those who have survived until eligibility (in black) are observed. Left truncation time, or the time between diagnosis and cohort entry, is shown with a dashed line.}
\label{fig:followup}
\end{figure}

Survival is defined as the time from diagnosis until death, but patients are not eligible to enter the cohort until a milestone is reached (black triangle). Patients 1 and 2 (in red) are excluded from the analysis because of death and right censoring, respectively, prior to study eligibility. A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors'' (those who have survived a long time, such as patients 3 and 7) and the recently-diagnosed.

Fortunately, the partial likelihood can be easily modified to make $R_i$ the set of individuals who are at risk (have entered the study and are not censored or failed) at that time assuming quasi-independence between left truncation and failure time \cite{tsai1987note}. Let individual $i$ have survival data of the form $(v_i, y_i, e_i, x_i)$, where individual $i$ enters the study at time $v_i$. The left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$. In other words, subjects are only considered at risk at a given event time after they have been observed to be at risk (i.e. exceeded their left truncation time, shown with a solid black line in \autoref{fig:followup} ). For example, at the time of death of Patient 3 in \autoref{fig:followup} , only Patient 4 has reached the milestone for study entry and is at risk. Patients 5, 6, and 7 reach their milestones at later time points.

@@

\subsection{Data generating process}
We generated $n$ patients and $p$ predictors to construct a $n \times p$ matrix, $X$. The first 11 predictors were from the CGDB and the remaining $\tilde{p}$ variables were simulated binary variables to represent a comprehensive set of CGDB binary alterations. The CGDB variables were the same $11$ variables used in the small $p$ model in \autoref{sec:rwd} and generated by randomly sampling rows of the original data with replacement. The matrix of binary variables, $\tilde{X}$ was simulated using a (latent) multivariate probit approach
